---
title: "Data 608 HW 3"
author: "Layla Quinones"
date: "10/21/2021"
output: 
  pdf_document: default
  html_document: default
runtime: shiny
---

```{r, message = FALSE, warning = FALSE}
library(shiny)
library(DT)
library(plotly)
library(ggplot2)
library(tidyverse)
library(googleVis)
library(rsconnect)
suppressPackageStartupMessages(library(googleVis))# hides messages
```

# Question 1

As a researcher, you frequently compare mortality rates from particular causes across different States. 
You need a visualization that will let you see (for 2010 only) the crude mortality rate, across all States, from one cause (for example, Neoplasms, which are effectively cancers). 

Create a visualization that allows you to rank States by crude mortality for each cause of death.

## Approach:

First we want to get started by installing all packages (above) and looking a the data to get a sense of what is in the raw data set to help inform data cleaning and next steps. 

```{r}
# Load data
rawData <- read.csv("https://raw.githubusercontent.com/charleyferrari/CUNY_DATA_608/master/module3/data/cleaned-cdc-mortality-1999-2010-2.csv", header = TRUE, stringsAsFactors = FALSE)

# look at the data
head(rawData)
```

We see that there is a column specifically for year so we want to subset the data by the year 2010.

```{r}
# Select 2010 only
data2010 <- subset(rawData, Year==2010)

#display new data
head(data2010)
```

```{r}
# Identify all possbile causes for death
causes <- unique(data2010$ICD.Chapter)

#display
causes
```


```{r}
# Create server 
UIone <- fluidPage(
  titlePanel("2010 CDC Mortality Rates"),
  fluidRow(selectInput("cause", "Cause of Death:", choices=sort(causes))), 
  plotOutput("plotOne", height = 800)
)

# Deine UI
serverOne <- function(input, output) {
  rates <- reactive({rates <- subset(data2010, ICD.Chapter==input$cause)}) #input must match the server
  output$plotOne <- renderPlot({ #output must match the server
    ggplot(rates(), aes(x = Crude.Rate, y = reorder(State, -Crude.Rate))) +
      scale_x_continuous(limits = c(0, max(rates()$Crude.Rate) + max(rates()$Crude.Rate)/10), expand = c(0,0)) + # allow axis to be a good size to show text
      expand_limits(x = 0) +  #force axis to start at 0
      geom_segment(aes(yend = State), xend = 0, color = "purple") +
      geom_point(size=3, color = "green") +
      geom_text(aes(x = rates()$Crude.Rate + max(rates()$Crude.Rate)/20, y = reorder(State, -Crude.Rate), label= rates()$Crude.Rate)) +
      theme_bw() +
      theme(panel.grid.major.y = element_blank(), axis.title=element_text(size=14))+
      xlab("2010 CDC Mortality Rate") +
      ylab("State") +
      ggtitle(input$cause)
  }) 
}
```


```{r}
#Run the shiny app
shinyApp(ui = UIone, server = serverOne)
```

# Question 2

Often you are asked whether particular States are improving their mortality rates (per cause) faster than, or slower than, the national average. Create a visualization that lets your clients see this for themselves for one cause of death at the time. Keep in mind that the national average should be weighted by the national population.

```{r}
# Get all states that appear at least once
states<-unique(rawData$State)

#display
states
```

```{r}
# Create UI script
UItwo <- fluidPage(
  title = "State Mortality Rates Over Time",
  fluidRow(
    column(6, selectInput('causes', 'Cause of Death', choices=sort(causes)) ), # allow user to choose state AND cause
    column(6, selectInput('states', 'State', choices=sort(states)) )
  ),
  fluidRow(
    plotOutput('myplot')
  )
)
```

```{r}
# Create server script
serverTwo <- function(input, output) {
  output$myplot <- renderPlot({
    
    #selecting only states and causes
    data <- rawData %>% 
      filter(State == input$states, ICD.Chapter == input$causes)
    
    #calculate the weighted average(considering the population) by Year
    usAverage <- rawData %>% 
      filter(ICD.Chapter == input$causes) %>% 
      group_by(Year) %>% 
      summarise(annualRate = (sum(as.numeric(Deaths))/sum(as.numeric(Population))*100000))
   
    #plot
    ggplot(data, aes(x=Year, y=Crude.Rate)) + 
      geom_point(aes(color = "blue"), size=2) + 
      geom_line(aes(x=Year, y=Crude.Rate, color = "blue"))+
      geom_line(aes(x = usAverage$Year, y = usAverage$annualRate, color='green'), size=2) + 
      scale_y_continuous(limits = c(0, max(usAverage$annualRate) + max(usAverage$annualRate/10), expand = c(0,0))) +
      expand_limits( y = 0) +  #force axis to start at 0
      geom_text(aes(x = usAverage$Year, y = usAverage$annualRate, label = round(usAverage$annualRate, digits = 1)))+
      geom_text(aes(x = Year, y = Crude.Rate, label = round(Crude.Rate, digits = 1)))+
      scale_color_manual(name='Legend', 
                         values=c('blue', 'green'),
                         labels=c('State', 'National Average'))
  })
}
```

```{r}
shinyApp(ui = UItwo, server = serverTwo)
```


